---
title: "Reading and writing data"
subtitle: "Getting data in and out of `Matlab`"
author: "Lincoln Colling, Reny Baykova"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    selt_contained: true
    nature:
      highlightStyle: github
      highlightLines: true 
      countIncrementalSlides: false
      beforeInit: "macros.js"
---

<!-- 
vim: set ft=markdown tw=80 spell spelllang=en_gb:
vim: set conceallevel=0 foldlevel=1:
-->

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, echo=FALSE, eval=TRUE}
library(xaringanthemer)

style_duo_accent(
 primary_color = "#003b49",
secondary_color = "#1d4289",
header_font_google = google_font("Cabin"),
text_font_google   = google_font("Noto Sans", "300", "300i"),
code_font_google   = google_font("Fira Mono"),
colors = c(
red = "#d3273e",
purple = "#5d3754s",
orange = "#dc582a",
green = "#007a78",
white = "#FFFFFF",
blue = "#1d4289"
)
)
xaringanthemer::style_extra_css(
  list(".title-slide" = list(
   # "background-image" =
   #   "url(https://upload.wikimedia.org/wikipedia/commons/thumb/3/34/University_of_Sussex_Logo.svg/480px-University_of_Sussex_Logo.svg.png)",
  "background-position" =  "95% 95%",
 "background-size" = "180px",
  "border" = "10px solid #013035",
  "background-color" = "#FFFFFF"
  ),
 ".title-slide h1" = list(
  "padding-top" = "0px",
  "font-size" = "60px",
  "text-align" = "left",
  "padding-bottom" = "18px",
  "margin-bottom" = "18px",
  "margin-top" = "0px",
  "color" = "#003b49"
 ),
 ".title-slide h2" = list(
   "font-size" = "40px",
   "text-align" = "left",
   "padding-top" = "10px",
   "margin-top" = "0px",
   "color" = "#003b49"
 ),
 ".title-slide h3" = list(
   "font-size" = "30px",
   "color" = "#26272A",
   "text-align" = "left",
   "text-shadow" = "none",
   "padding" = "0px",
   "margin" = "0px",
   "line-height" = "1"),
   ".midpage h3" = list(
   "font-size" = ".6em",
   "margin-bottom" = ".2em"),
 ".footnote" = list(
   "font-size" = ".6em"
 ),
   ".section h1" = list(
    "padding" = "auto",
    "display" = "block",
    "margin-left" = "auto",
    "margin-right" = "auto",
    "padding-top" = "250px",
    "vertical-align" = "auto",
    "align-items" = "center",
    "font-size" = "40px"
   ),
".section .remark-slide-number" = list(
   "display" = "none"),
   "pre" = list("border-style" = "solid",
   "border-width" = "1px",
   "border-color" = "lightgrey",
   "padding" = "5px",
   "border-radius" = "3px")
))

```


```{r broadcast, echo=FALSE}
xaringanExtra::use_broadcast()
xaringanExtra::use_share_again()
xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons"))
```


```{r, echo=FALSE, eval=FALSE}
# I'll add some meta data later
# library(metathis)
# meta() %>%
#   meta_description(desc)
```

# Data formats

When it comes to reading/writing data in `Matlab`, there are a few options.

**Matlab-specific formats**  

- `.mat` files

**Domain-specific formats**  

- Data formats specific to certain applications: e.g., EEG data, fMRI data, Eye
  tracking data

- Cognitive neuroscience science-specific formats: e.g., BIDS (Brain Imaging
  Data Structure)


**General data formats**

- `.csv` (comma-separated values)

- `.json` (JavaScript Object Notation)

The different formats have different advantages and disadvantages. Sometimes
you'll have a choice in the format you use, and sometimes you won't. 

---

# `Matlab` specific format

The easiest format to use in `Matlab` is the `.mat` format

```matlab
`r paste0(xfun::read_utf8("../../code/part3/save_data.m"), collapse="\n")`
```

With the `save` command, you can quickly and easily save the entire workspace.


```matlab
`r paste0(xfun::read_utf8("../../code/part3/load_data.m"), collapse="\n")`
```

And with the `load` command, you can quickly load a `.mat` file into the
workspace.

---

## `.mat` files

Using `save` and `load` to save and load `.mat` files is quick and easy, but
there are a few downsides 

- Saves the **entire** workspace, which could be **HUGE** (many gigabytes if
  you're working with imaging data)

- Loads **all** the saved data at once (again, possibly **HUGE**), which can
  overwrite variables (without warning) if you have naming conflicts.

- It's not easy to work with `.mat` files in other languages (e.g., it's
  **extremely** slow to open them in `R`—on the order of minutes, even for
  relatively small files).

---

## Selectively saving and loading

Instead of saving the entire workspace, we can use the `save` function to save
particular variables. 

You just need to list the **names** of the variables you want to save.

**Saving specific variables**  

```matlab
`r paste0(xfun::read_utf8("../../code/part3/create_many_save_few.m"), collapse="\n")`
```

---

## Selectively saving and loading

You can also selectively load variables in a `.mat` file.

Again, you just need to list the **names** of the variables you want to load.

**Loading specific variables**  

```matlab
`r paste0(xfun::read_utf8("../../code/part3/load_few.m"), collapse="\n")`
```

Alternately, in case there are naming conflicts, you can load the content of a
`.mat` file into a variable

**Loading into a variable**  

```matlab
`r paste0(xfun::read_utf8("../../code/part3/load_into_var.m"), collapse="\n")`
```

---


## Domain-specific formats

**Typically**, you'll be working with domain-specific formats.

- Formats specific to EEGLab or FieldTrip 

- NIfTI/DICOM for MRI data

However, there's also a generic format for storing all stored to cognitive
neuroscience data called [BIDS](https://bids.neuroimaging.io)

- MRI Data

- M/EEG

- iEEG

- Behavioural data

- Genetic data

I think the usage of the BIDS format will become more and more common because it
doesn't lock you into any specific programming platform (e.g., `Matlab` or
`Python`) or any specific analysis toolbox (e.g., SPM, FieldTrip, EEGlab)

---

## General data formats

Finally, there are a few general-purpose data formats that you'll be familiar
with from using, e.g., `R` and `Python`.

**CSV files**  

- `.csv` files are good for storing matrices of numbers

- They're easy to import into `R`, `Python` and `SPSS`

However, matrices in `Matlab` don't have column headers (e.g., they're not
tables) so column headers will be missing when you read the data in to `R`

`Matlab` has some build in functions for reading data from `.csv` files into
matrices and for writing matrices into `.csv` files that we'll use in the
activities, but we'll also look at some alternatives.


---

**Tips for working with `.csv` files**

- Check out the `Matlab` **file exchange** for [a simple
  function](https://uk.mathworks.com/matlabcentral/fileexchange/29933-csv-with-column-headers)
  for writing `.csv` files with headers. You'll need a matrix that holds the
  data and another cell array that holds the header

- Convert the matrix to a **table** (tables are `Matlab`'s version of a
  data.frame).

- Or organise your data as a `struct` array and use the `writeStruct2csv`
  function included in the notes. I think this is a good approach for working
  with behavioural data from PsychToolbox because the resulting data will
  be easy to analyse in `R`, `Python`, or SPSS.


---


Sometimes we want to store more complex data:

A typical use case for this might be for saving `struct`s

Here, we can see an example of creating a `struct` for storing some data about
a participant: 

```matlab
SubjectInfo = struct;
SubjectInfo.age = 19
SubjectInfo.Handedness = 'Left'
```

Because this data isn't matrix (or at least some kind of table) it's not
suitable for storing in a `.csv` file. 

We need an alternative. One alternative is to use `.json` files.

**JSON files**

- `.json` files are good for storing non-matrix data

- They're easy to import into `R` and `Python`

To save data as a `.json` file, you first need to **encode** it, and then you
can save it.

---

**Working with JSON files**

The first step to save data as a `.json` file is to encode the variable as JSON.

You can then write the data to a file using `fopen`, `fprintf`, and `fclose`.

Here is an example:
```matlab
JSONFile= 'example.json'
SubjectInfo_json = jsonencode(SubjectInfo);
fid = fopen(JSONFile, 'w');
fprintf(fid, SubjectInfo_json);
fclose(fid);
```

This is a good example of something you might want to write a function for! This
is what it might look like:


```matlab
function writeJSON(filename, variable)
  encoded_variable = jsonencode(variable);
  fid = fopen(filename, 'w');
  fprintf(fid, encoded_variable);
  fclose(fid);
end

```

---

### Working with JSON files in other languages

The big advantage of `.json` files is that they allow us to import structured
data into other programming languages

**Reading a JSON file into `R`**

```r
jsonlite::fromJSON('example.json')
```

**Reading a JSON file into `Python`**

```python
import json
with open('example.json') as f:
  data = json.load(f)
```
