---
title: "Flow control and loops"
subtitle: "Controlling program flow and looping"
author: "Lincoln Colling, Reny Baykova"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    selt_contained: true
    nature:
      highlightStyle: github
      highlightLines: true 
      countIncrementalSlides: false
      beforeInit: "macros.js"
---

<!-- 
vim: set ft=markdown tw=80 spell spelllang=en_gb:
vim: set conceallevel=0 foldlevel=1:
-->

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, echo=FALSE, eval=TRUE}
library(xaringanthemer)

style_duo_accent(
 primary_color = "#003b49",
secondary_color = "#1d4289",
header_font_google = google_font("Cabin"),
text_font_google   = google_font("Noto Sans", "300", "300i"),
code_font_google   = google_font("Fira Mono"),
colors = c(
red = "#d3273e",
purple = "#5d3754s",
orange = "#dc582a",
green = "#007a78",
white = "#FFFFFF",
blue = "#1d4289"
)
)
xaringanthemer::style_extra_css(
  list(".title-slide" = list(
   # "background-image" =
   #   "url(https://upload.wikimedia.org/wikipedia/commons/thumb/3/34/University_of_Sussex_Logo.svg/480px-University_of_Sussex_Logo.svg.png)",
  "background-position" =  "95% 95%",
 "background-size" = "180px",
  "border" = "10px solid #013035",
  "background-color" = "#FFFFFF"
  ),
 ".title-slide h1" = list(
  "padding-top" = "0px",
  "font-size" = "60px",
  "text-align" = "left",
  "padding-bottom" = "18px",
  "margin-bottom" = "18px",
  "margin-top" = "0px",
  "color" = "#003b49"
 ),
 ".title-slide h2" = list(
   "font-size" = "40px",
   "text-align" = "left",
   "padding-top" = "10px",
   "margin-top" = "0px",
   "color" = "#003b49"
 ),
 ".title-slide h3" = list(
   "font-size" = "30px",
   "color" = "#26272A",
   "text-align" = "left",
   "text-shadow" = "none",
   "padding" = "0px",
   "margin" = "0px",
   "line-height" = "1"),
 ".footnote" = list(
   "font-size" = ".6em"
 ),
   ".section h1" = list(
    "padding" = "auto",
    "display" = "block",
    "margin-left" = "auto",
    "margin-right" = "auto",
    "padding-top" = "250px",
    "vertical-align" = "auto",
    "align-items" = "center",
    "font-size" = "40px"
   ),
".section .remark-slide-number" = list(
   "display" = "none"),
   "pre" = list("border-style" = "solid",
   "border-width" = "1px",
   "border-color" = "lightgrey",
   "padding" = "5px",
   "border-radius" = "3px"),
 "larger" = list("font-size" = "2em")))
```


```{r broadcast, echo=FALSE}
xaringanExtra::use_broadcast()
xaringanExtra::use_share_again()
xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons"))
```


```{r, echo=FALSE, eval=FALSE}
# I'll add some meta data later
# library(metathis)
# meta() %>%
#   meta_description(desc)
```

# <small>Flow Control</small>

It's common to want to write a program (a script or a function) that will do
something different depending on whether some condition is met. That is, 
an **IF A THEN B** rule.

- In `Matlab`, we can do this with an `if` statement

```matlab
birthyear = 1992
if birthyear <  2002
    disp('You are old enough to buy beer')
end % notice that we put end when we've finished the rule
```

If the condition `birthyear < 2002` produces a logical `1`, then the path is
followed:

```matlab
>> birthyear = 1992;
>> birthyear < 2002

ans =

  logical

   1
```

---

## `if` then `else`

You can also use `else` to specify what to do if the condition _isn't_ met:

```matlab
if birthyear < 2002
    disp('You are old enough to buy beer')
else
    disp('You are not old enough to buy beer!')
end
```

Or `elseif` for more complex branching:

```matlab
if birthyear < 2002
    disp('You are old enough to buy beer')
elseif birthyear == 2002 
    disp('You might be old enough to buy beer')
else 
    disp('You are not old enough to buy beer!')
end
```

---

## `switch` and `case`

If you need to check through **many** conditions, then there's a more efficient
way than stringing together a bunch of `if`, and `elseif` statements.

Instead, you can use `switch` and `case`:

```matlab
n = 10 % set n to a number

switch n 
  case 1
      disp('one')
  case 2
      disp('two')
  otherwise 
      disp('I can only count to two')
end
```

---


## `switch` and `case`

`switch` and `case` also works with strings  and `chars`

```matlab
n = 'one'

switch n
  case 'one'
      disp('1')
  case 'two'
      disp('2')
  case {'three','four'}
      disp('more than 2 less than 5')
end
```

---

# Loops

Loops allow us to run a bit of code over and over.

- Loops are extremely useful in, e.g., PsychToolbox where we present multiple trials,
  or in data analysis where we might want to process multiple data files.

`Matlab` has **two** basic forms of loops:

- The `for` loop, which loops a certain number of times.

- The `while` loop, which loops continuously as long as some condition is met.

Whether you use a `while` loop or `for` loop will depend on the **problem you
want to solve** or how you **structure your solution**. 

---

## The `for` loop

- The simplest **loop** is the `for` loop.

- If we pass a range of values (e.g., a vector or cell array) to a `for`
  statement, then one value of that range will be available on each iteration of
  the loop

An example with a vector of numbers:

```matlab
for i = 1 : 10
    disp(num2str(i))
end
```

An example with a cell array:

```matlab
for n = {'one','two','three'}
    disp(n)
end
```

---

## The `while` loop

- `for` loops loop over some series of values.

- `while` loops, on the other hand, loop continuously as long as some condition
  is met.

```matlab
x = 1
y = 0
while y < 100
    y = x^2
    disp(['x squared is ' num2str(y)])
    x = x + 1
end 
```

This means `while` loops are useful when we don't know how long we have to loop
for.

```matlab
% read lines from a file
% carry on looping while you're not at the end of the file
while ~feof(fid)
    fgetl(fid)
end
```

<!--
/* --- */

## Alternatives to loops<sup>1</sup>

If you've used `map`-style functional syntax in `R` (either through `purrr::map()` or `apply()`) or `python` (either with `map()` or a list-comprehension with a lambda)

- In `Matlab` you can use `arrayfun` and `cellfun` to **apply** a function to every element of a 1-D matrix or 1-D cell array

```matlab
>> list_of_numbers = [1 6 8 12 12];
>> logs = arrayfun(@(x) log(x), list_of_numbers)

logs =

         0    1.7918    2.0794    2.4849    2.4849
```


.footnote[<sup>1</sup>Knowing this isn't vitally important, but if you like using functional style programming in `R` or `python` then you might want to do the same in `Matlab`]

-->
