---
title: "Working with data"
subtitle: "Data types, indexing, and logic"
author: "Lincoln Colling, Reny Baykova"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    selt_contained: true
    nature:
      highlightStyle: github
      highlightLines: true 
      countIncrementalSlides: false
      beforeInit: "macros.js"
---

<!-- 
vim: set ft=markdown tw=80 spell spelllang=en_gb:
vim: set conceallevel=0 foldlevel=1:
-->

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, echo=FALSE, eval=TRUE}
library(xaringanthemer)

style_duo_accent(
 primary_color = "#003b49",
secondary_color = "#1d4289",
header_font_google = google_font("Cabin"),
text_font_google   = google_font("Noto Sans", "300", "300i"),
code_font_google   = google_font("Fira Mono"),
colors = c(
red = "#d3273e",
purple = "#5d3754s",
orange = "#dc582a",
green = "#007a78",
white = "#FFFFFF",
blue = "#1d4289"
)
)
xaringanthemer::style_extra_css(
  list(".title-slide" = list(
   # "background-image" =
   #   "url(https://upload.wikimedia.org/wikipedia/commons/thumb/3/34/University_of_Sussex_Logo.svg/480px-University_of_Sussex_Logo.svg.png)",
  "background-position" =  "95% 95%",
 "background-size" = "180px",
  "border" = "10px solid #013035",
  "background-color" = "#FFFFFF"
  ),
 ".title-slide h1" = list(
  "padding-top" = "0px",
  "font-size" = "60px",
  "text-align" = "left",
  "padding-bottom" = "18px",
  "margin-bottom" = "18px",
  "margin-top" = "0px",
  "color" = "#003b49"
 ),
 ".title-slide h2" = list(
   "font-size" = "40px",
   "text-align" = "left",
   "padding-top" = "10px",
   "margin-top" = "0px",
   "color" = "#003b49"
 ),
 ".title-slide h3" = list(
   "font-size" = "30px",
   "color" = "#26272A",
   "text-align" = "left",
   "text-shadow" = "none",
   "padding" = "0px",
   "margin" = "0px",
   "line-height" = "1"),
 ".footnote" = list(
   "font-size" = ".6em"
 ),
   ".section h1" = list(
    "padding" = "auto",
    "display" = "block",
    "margin-left" = "auto",
    "margin-right" = "auto",
    "padding-top" = "250px",
    "vertical-align" = "auto",
    "align-items" = "center",
    "font-size" = "40px"
   ),
".section .remark-slide-number" = list(
   "display" = "none"),
   "pre" = list("border-style" = "solid",
   "border-width" = "1px",
   "border-color" = "lightgrey",
   "padding" = "5px",
   "border-radius" = "3px")
))

```


```{r broadcast, echo=FALSE}
xaringanExtra::use_broadcast()
xaringanExtra::use_share_again()
xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons"))
```

# Variables and data types

- Variables are used for storing information.

- You can _assign_ content to a variable with an `=` sign.

- There are different **types** of data that we can assign to a variable.


We've already had the chance to play around with variables a little, so some of
this should be familiar by now.

---


## <smaller>Matlab's data types</smaller>

The primary data types that we'll be working with in `Matlab` are:

1. Numeric arrays (matrices) - for storing numbers 

2. Character arrays and string arrays  - for storing text

3. Cell arrays - for storing mixed types of data

4. Structures arrays - for storing mixed types of data that you can refer to by name


.footnote[`Matlab` also has a data type called a table (which is similar to a
`tibble` or `data.frame` in `R`), but I won't be talking about them much.]

---

### Numeric arrays/ matrices

The base data type in `Matlab` is the **matrix** and it stores numeric data. 

A matrix in `Matlab` can have one dimension (*vector*), two dimensions (*matrix*), or more but in `Matlab`, we refer to them all as a *matrix*.

The simplest matrix in `Matlab` is a 1 × 1 matrix. A 1 × 1 matrix is just a single number:

```matlab
>> a_number = 1;
```

A 1 × 4 matrix would be a list of four numbers.

To create a 1 x 4 matrix, we wrap a list of four numbers in `[]` and separate them with `,` or spaces:

```matlab
>> a_vector_commas = [1, 2, 3, 4];

>> a_vector_spaces = [1 2 3 4]

a_vector_spaces =

     1     2     3     4
```

---

### Numeric arrays/ matrices

We could also create, e.g., a 4 × 1 matrix, which would be a list of four
numbers arranged vertically instead of horizontally.

To do this, we separate the numbers with `;` instead of `,` or spaces:

```matlab
>> a_column_vector = [1; 2; 3; 4]

a_column_vector =

     1
     2
     3
     4

```

<small>The distinction between row vectors (1 × 4) and column vectors (4 × 1) is
not usually that important, but it does become important when we
want to *concatenate* two vectors together.</small>

---

#### 2-D matrices

More commonly we'll encounter matrices that have multiple rows and columns
(e.g., a 2 × 3 matrix).

When creating a 2-D matrix, we use `,` or spaces to create separate columns, and we use `;` to create separate rows:

```matlab
>> a_matrix = [11, 12, 13; 21, 22, 23] % create a 2 x 3 matrix

a_matrix =

    11    12    13
    21    22    23
```

---

#### 2-D matrices

Matrices in Matlab must be **rectangular**:

```matlab
>> a_matrix = [11, 12, 13; 21, 22, 23] % a valid matrix

a_matrix =

    11    12    13
    21    22    23
```

Otherwise, you'll get an error:

```matlab
>> a_matrix = [11, 12; 21, 22, 23] % an invalid matrix
Error using vertcat
Dimensions of arrays being concatenated are not consistent.
```

---

### Accessing and indexing matrices

Once we have a matrix of data, we might want to access specific elements in that
matrix.

To access a single element of a matrix we access using the following syntax:

```matlab
matrix_name(row_number, column_number)
```

For example: 

```matlab
a_matrix(1,1); % get row 1 column 1

a_matrix(2,1); % get row 2 column 1

a_matrix(2,2); % get row 2 column 2
```

---

#### Accessing multiple elements 

We can also access more than one element at a time. For example, we can get entire
rows or columns. To do this we use `:` to get all the rows/columns:

```matlab
a_matrix(1,:); % get the first row and all columns

a_matrix(:,1); % get the first column and all the rows
```

We can also access  multiple **specific** elements, not just the whole row or column:

```matlab
a_matrix(1,1:3); % get the first row, columns 1, 2, and 3

a_matrix(1,[1,3]); % get the first row, columns 1 and 3
```

---

#### Access short-cuts for 1-D matrices

There are a few shortcuts we can use for accessing elements in 1-D matrices.  

For example, if we have a 1 × N or N × 1 matrix then we can use
a single index value:

```matlab
a_row_vector = [1, 2, 3, 4];
a_col_vector = [1; 2; 3; 4];

a_row_vector(2) % the second element of the vector
% this is the same as:
a_row_vector(1,2) 

a_col_vector(4) % the fourth element of the vector
% this is the same as:
a_col_vector(4,1)
```

---

#### The `end` keyword 

`Matlab` also has a special keyword called `end` which we can use to access the
last row/column/element of a matrix:

```matlab
a_row_vector(end) % the last element of the row vector
a_row_vector(end-1) % the second to last element of the row vector

a_matrix(end,:) % the last row of the matrix

a_matrix(:,end) % the last column of the matrix
```

---

#### Logical indexing of matrices

We can use logical indexing to access elements (see [the cheat
sheet](/cheatsheet.html#logical-operations-numbers) for examples of logical
rules).

If we use one of the logical operators together with a matrix or a vector we
get an output that tells us which elements match our rule:

```matlab
>> a_matrix >= 22

ans =

  2x3 logical array

   0   0   0
   0   1   1
```

We use this logical matrix output to index the elements we want:

```matlab
>> a_matrix(a_matrix >= 22)

ans =

    22
    23
```

---

#### Modifying matrices

Instead of *accessing* elements of a matrix, we can also *modify* a matrix
*assigning* new values to its elements.

First, our original vector:
```matlab
>> a_vector 

a_vector =

     1     2     3     4     5
```

Then we change the last element to `99`:
```matlab
>> a_vector(end) = 99;
```

Now the last element has been changed to `99`:
```matlab
>> a_vector

a_vector =

     1     2     3     4    99
```

---

#### Modifying matrices 

We can also use logical rules to decide which elements to modify.

For example, we could change all the elements in `a_matrix` that are *greater
than 21* to `NaN` (not a number):

```matlab
>> a_matrix(a_matrix > 21) = NaN 

a_matrix =

    11    12    13
    21   NaN   NaN

```

In `Matlab`, if you try to insert a value in a new position, then `0`s are
added to fill any intermediate positions<sup>1</sup>:

```matlab
>> a_vector = [1, 2, 3]; % matrix has 3 elements
>> a_vector(11) = 99 % modifying the 11th element

a_vector =

     1     2     3     0     0     0     0     0     0     0    99
```

.footnote[<sup>1</sup>This is in contrast to `R` which will fill the
intermediate positions with `NA` or `Python` which will produce an error.]

---

### Character arrays and string arrays

Matrices can only store numbers. But sometimes we want to store text.

`Matlab` provides two ways of storing text.

First: A `char` is a sequence of characters, just as a vector is a sequence of numbers. Each character in a `char array` is a unique element.

To assign a series of letters to a variable as a `char` we put the letters
inside single quotes `' '`:

```matlab
>> a_char = 'A char of letters';
>> length(a_char) % check the numbers of elements in the array

ans =

    17
```

<u>Most of the time you'll want to use `char`s to store characters.</u>

---

### Character arrays and string arrays

Second: A `string` allows you to store a sequence of characters as a single element, and a `string array` allows you to store multiple strings (each being a separate element).

To store letters as a `string` we put the letters inside double quotes `"
"`<sup>1</sup>:

```matlab
>> a_string = "A string of letters"; % a string array with 1 element
>> length(a_string) % check the numbers of elements in the array

ans =

     1
     
>> my_name = ["Reny", "Baykova"]; % a string array with 2 elements
>> length(my_name)

ans =

     2
     
```

Note that `Matlab` **does** distinguish between `''` and
`""` whereas, for example, `R` does not make a distinction. Strings are a relatively new data type in `Matlab` and older Matlab versions
would produce an error if you used `""`.

---

####  Concatenating with chars/strings

You'll almost always want to use `char`s, which means that you'll almost always
use single quotes `''`.


`Char`s and `String`s behave differently in `Matlab` when
concatenating:

```matlab
>> first_name_char = 'Reny';
>> last_name_char = 'Baykova';
>> full_name_char = [first_name_char, last_name_char]

full_name_char =

    'RenyBaykova'
```

```matlab
>> first_name_str = "Reny";
>> last_name_str = "Baykova";
>> full_name_str = [first_name_str, last_name_str]

full_name_str = 

  1x2 string array

    "Reny"    "Baykova"
```

---


####  Indexing with chars/strings

Indexing using () also works differently. With a `char`, we can access individual *characters* because each character is a separate element:

```matlab
>> full_name_char

full_name_char =
    'RenyBaykova'

>> full_name_char(1)

ans =
    'R'
```

In contrast, each element in a `string` can be a sequence of characters:

```matlab
>> full_name_str

full_name_str = 
  1×2 string array

    "Reny"    "Baykova"
    
>> full_name_str(1)

ans = 
    "Reny"    
```

---

#### Logic with strings

Logical operations also behave differently between `char`s and `string`s.

And the behaviour with `char`s (which is what you'll mostly be using) may **not**
be what you'd expect.

First, let's compare two `string`s:

```octave
>> "ab" == "ab" 

ans =

  logical

   1
```

```octave
>> "ab" == "ac"

ans =

  logical

   0
```

This is the behaviour you'd probably expect. 


---


#### Logic with chars

Now, let's compare two `char` arrays:

```matlab
>> 'ab' == 'ab'

ans =

  1x2 logical array

   1   1
```

```matlab
>> 'ab' == 'ac'

ans =

  1x2 logical array

   1   0
```

Logical operations on `char` arrays operate on each character.

Which most of the time is **not** what we want!

---

#### Logic with string/chars

Instead of using `==` to compare the two arrays element-by-element... 
Instead of logical equality (`==`), we can use `strcmp()` to compare `char`s:

```matlab
>> 'ab' == 'ac'

ans =

  1x2 logical array

   1   0

```

We can use `strcmp()`:

```matlab
>> char_one = 'ab';
>> char_two = 'ac';
>> strcmp(char_one, char_two)

ans =

  logical

   0
```

`strcmp()` will compare the entire string of text… so we'll mostly use
`strcmp()` for logic with `string`s and `char`s 


---

class: section

# Advanced data types 

Cell arrays and structure arrays


---

### Cell arrays

We've covered matrices and string/char arrays for when we want to store numeric data and
text, respectively. 

But sometimes we want to store text and numeric data in one variable.

`Matlab` provides two additional data types for doing this: `cell` arrays and
`struct` arrays.

Cell arrays are *a bit* like matrices, but we use `{}` instead of `[]` when
creating them:

```matlab
>> a_cell_array = {'a letter', 99; "a", [12 ; 12]}

a_cell_array =

  2x2 cell array

    {'a letter'}    {[      99]}
    {["a"     ]}    {2x1 double}

```

Here we have a 2 × 2 cell array where one element is a char, another is a 
one-element matrix, another one is a string, and the final one is a column vector.

Each cell in a _cell array_ can hold any other kind of `Matlab` data.


---

#### Accessing and indexing cell arrays

With matrices, there's no distinction between an element of a matrix and the
*content* of that element, but there is this distinction for cell arrays.

```matlab
a_cell_array = {'element 1', [2; 3]}; % first we create a cell array
```

If we want to get the *element* then we use `()`, just like with matrices:

```matlab
>> a_cell_array(2)

ans =

  1x1 cell array

    {2x1 double}
```

Using `()` returns a 1 × 1 cell array that contains a 2 x 1 matrix, but we can't see what that matrix contains.

---

#### Accessing and indexing cell arrays

If we want the *content* (i.e., the matrix that is in cell 2), then we can
use `{}` instead:

```matlab
>> a_cell_array{2}

ans =

     2
     3
```

Using `{}` returns the 2 × 1 matrix in cell 2 of the cell array. 

---

### Structure arrays 

- Like `cell` arrays, `struct` arrays allow you to mix different data types. 

- `struct` arrays organise data in key-value pairs<sup>1</sup>.
The data goes in different **fields** and the name you want to assign to each field is stored in the associated **field name**.

- `struct` arrays are a common way of organising data in, for example, SPM, EEGLab,
  and FieldTrip.

To create an empty `struct` array we can use the `struct` function.

```matlab
person = struct % create an empty struct named person

person.name = 'Reny' % we create field called "name", and assign a char to it

person.pets = {'dog' 'cat'} % assign a cell array to "pets" field

person.age = [32]; % assign a matrix to the "age" field

```


.footnote[<sup>1</sup>A `struct` is a like a named list in `R` or a `dict` in
`Python`. They're also a way of organising data that is similar to `.json`
files.]

---

#### Accessing structure arrays

Once we have a `struct` array like this:

```matlab
>> person

person = 

  struct with fields:

    name: 'Reny'
    pets: {'dog'  'cat'}
     age: 32
```

…then we can use the field names to access the elements:

```matlab
>> person.name

ans =

    'Reny'
```

---

#### Accessing structure arrays

We can also use variables to *dynamically* access `structs`. 

First, let's get the names of the fields using `fieldnames`:
```matlab
>> the_fields = fieldnames(person)

the_fields =

  3x1 cell array

    {'name'}
    {'pets'}
    {'age' }

>> the_field_I_want = the_fields{1}

the_field_I_want =

    'name'
```

Now I have a variable called `the_field_I_want` that contains a `char`
corresponding to the name of a field.

---


#### Accessing structure arrays

I can use the variable `the_field_I_want` (or any other variable that
happens to hold a char corresponding to a field name) to access that fiel.


To do this, I use the variable name inside `()` instead of using the
field name itself:

```matlab
>> person.(the_field_I_want)

ans =

    'Reny'
```

The use case for this might not be immediately obvious, but we'll run through
some scenarios in later problem-solving sections.

