---
title: "Functions"
subtitle: "Reusable code using functions"
author: "Lincoln Colling, Reny Baykova"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    selt_contained: true
    nature:
      highlightStyle: github
      highlightLines: true 
      countIncrementalSlides: false
      beforeInit: "macros.js"
---

<!-- 
vim: set ft=markdown tw=80 spell spelllang=en_gb:
vim: set conceallevel=0 foldlevel=1:
-->

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, echo=FALSE, eval=TRUE}
library(xaringanthemer)

style_duo_accent(
 primary_color = "#003b49",
secondary_color = "#1d4289",
header_font_google = google_font("Cabin"),
text_font_google   = google_font("Noto Sans", "300", "300i"),
code_font_google   = google_font("Fira Mono"),
colors = c(
red = "#d3273e",
purple = "#5d3754s",
orange = "#dc582a",
green = "#007a78",
white = "#FFFFFF",
blue = "#1d4289"
)
)
xaringanthemer::style_extra_css(
  list(".title-slide" = list(
   # "background-image" =
   #   "url(https://upload.wikimedia.org/wikipedia/commons/thumb/3/34/University_of_Sussex_Logo.svg/480px-University_of_Sussex_Logo.svg.png)",
  "background-position" =  "95% 95%",
 "background-size" = "180px",
  "border" = "10px solid #013035",
  "background-color" = "#FFFFFF"
  ),
 ".title-slide h1" = list(
  "padding-top" = "0px",
  "font-size" = "60px",
  "text-align" = "left",
  "padding-bottom" = "18px",
  "margin-bottom" = "18px",
  "margin-top" = "0px",
  "color" = "#003b49"
 ),
 ".title-slide h2" = list(
   "font-size" = "40px",
   "text-align" = "left",
   "padding-top" = "10px",
   "margin-top" = "0px",
   "color" = "#003b49"
 ),
 ".title-slide h3" = list(
   "font-size" = "30px",
   "color" = "#26272A",
   "text-align" = "left",
   "text-shadow" = "none",
   "padding" = "0px",
   "margin" = "0px",
   "line-height" = "1"),
 ".footnote" = list(
   "font-size" = ".6em"
 ),
   ".section h1" = list(
    "padding" = "auto",
    "display" = "block",
    "margin-left" = "auto",
    "margin-right" = "auto",
    "padding-top" = "250px",
    "vertical-align" = "auto",
    "align-items" = "center",
    "font-size" = "40px"
   ),
".section .remark-slide-number" = list(
   "display" = "none"),
   "pre" = list("border-style" = "solid",
   "border-width" = "1px",
   "border-color" = "lightgrey",
   "padding" = "5px",
   "border-radius" = "3px"),
 "larger" = list("font-size" = "2em")))
```


```{r broadcast, echo=FALSE}
xaringanExtra::use_broadcast()
xaringanExtra::use_share_again()
xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons"))
```


```{r, echo=FALSE, eval=FALSE}
# I'll add some meta data later
# library(metathis)
# meta() %>%
#   meta_description(desc)
```

# Functions 

So far, we've been using functions that come with `Matlab`, but often we'll
want to write our own functions.

## <small>What are functions?</small> 

- Functions work like *mathematical functions*. They transform
  **inputs**<sup>1</sup> to **outputs**.

- They’re **self-contained bits** of code with rules for transforming the
  information.

- Some examples of functions include the `help`, `doc`, `vectcat`, `horzcat`,
  `randn`, and `strcmp` functions from [Part I](../part1.html).

.footnote[<sup>1</sup>*Technically*, functions don't need to take inputs or
produce outputs, but we'll get to that in a while.]

---

## Using functions 

Functions in `Matlab` can take between 0 to *n* inputs and can produce between 0
and *n* outputs. 

When you specify multiple **inputs** for functions, then they go inside `()`
after the function name, and are separated by `,`.

When you specify multiple **outputs** for functions, then they go inside `[]`
before the `=`, and are separated by `,`.

```matlab
>> x = rand(10, 1); % rand with two inputs and one output
>> y = rand(10, 1); % rand with two inputs and one output
>> [rho, pval] = corr(x, y) % corr with two inputs and two outputs

rho =

    0.2682


pval =

    0.4537
```

---

## Why write functions?

One principle that should guide the code you write can be summed up as:

<larger><u><strong>D</strong></u>on't <u><strong>R</strong></u>epeat
<u><strong>Y</strong></u>ourself</larger>

If you find yourself **repeatedly** using a bit of code (i.e., more than twice)
then you should probably write a function.

Functions allow you to **encapsulate** a bit of code

- The code **inside** a function exists separately from the rest of the code in
  a script. 

- This means that functions will **always** produce the **same output** if
  they're given the **same input**.  

---

## Our first function

Let's write a function to work out the **mean** of some numbers:

```matlab
% work out the mean of some numbers 
function mean_value = my_mean(input_numbers)

  % we'll use length to get the number of items
  length_of_input = length(input_numbers)

  % now we'll use sum to add up all the items
  sum_of_input = sum(input_numbers)

  % now work out the mean!
  mean_value = sum_of_input / length_of_input

end

```
---

### Taking apart our first function

Our function starts with a comment. This is what is shown when we use `help`.

```matlab
* % work out the mean of some numbers 
function mean_value = my_mean(input_numbers)

  % we'll use length to get the number of items
  length_of_input = length(input_numbers)

  % now we'll use sum to add up all the items
  sum_of_input = sum(input_numbers)

  % now work out the mean!
  mean_value = sum_of_input / length_of_input

end

```

Check the **help**…
```matlab
>> help my_mean
  work out the mean of some numbers
```

---

### Taking apart our first function

Next, we have the **function declaration** which includes the names of the
**output variables**, the names of the **input variables**, and the **name of
the function**.

```matlab
% work out the mean of some numbers 
*function mean_value = my_mean(input_numbers)

  % we'll use length to get the number of items
  length_of_input = length(input_numbers)

  % now we'll use sum to add up all the items
  sum_of_input = sum(input_numbers)

  % now work out the mean!
  mean_value = sum_of_input / length_of_input

end

```

---

### Taking apart our first function

Then we have the code that does the actual work…

```matlab
% work out the mean of some numbers 
function mean_value = my_mean(input_numbers)

* % we'll use length to get the number of items
* length_of_input = length(input_numbers)
*
* % now we'll use sum to add up all the items
* sum_of_input = sum(input_numbers)
*
* % now work out the mean!
* mean_value = sum_of_input / length_of_input

end

```

Note that this part of the function can only include references to variables
that were listed as **inputs** or variables that were **created in the
function** itself

We'll talk more about this when we cover **variable scope**.

---

### Taking apart our first function

Finally, we have an `end` to indicate that this is the **end** of the function.

```matlab
% work out the mean of some numbers 
function mean_value = my_mean(input_numbers)

  % we'll use length to get the number of items
  length_of_input = length(input_numbers)

  % now we'll use sum to add up all the items
  sum_of_input = sum(input_numbers)

  % now work out the mean!
  mean_value = sum_of_input / length_of_input

*end

```

---

#### Header comments 

The comments at the start (before the function declaration) are shown when you
use the `help` command.

Use these to write help for your functions.

Things to include in the help:

- What the function does

- What inputs it needs

- What outputs it produces

```matlab
% say_hello
% Function says hello to person 
% 
% Usage: 
% done = say_hello(input_name)
%
% Inputs:
% input_name : a char with a person's name
%
% Outputs:
% done : logical value that indicates that the function was run
```

---

#### The function declaration 

The **function declaration** indicates that it's a function, and it specifies the
**name** of the function.

It also specifies any **inputs** the function needs or any **outputs** the
function produces.

```matlab
function [ouput1, output2] = function_name(input1, input2, input3)
```

- In the example above, values 1–3 will be accessible in variables called
  `input1`, `input2`, and `input3`

- Whatever values are assigned to `output1` and `output2` by the time the
  function finishes running will be produced as outputs<sup>1</sup> 

- The function with `function_name` should be stored in a file called
  `function_name.m`

.footnote[<sup>1</sup>Note, `Matlab` does not use a `return` statement like most
other languages]

---

#### The function body

The **body of the function** is where you put the code that does the
work.

- Only values passed in as **inputs** or variables **created inside** the
  function body will be accessible.

- You can't access variables in the **workspace** or write to variables in the
  **workspace**

```matlab
function mean_value = my_mean(input_numbers)

  % we'll use length to get the number of items
  length_of_input = length(input_numbers)

  % now we'll use sum to add up all the items
  sum_of_input = sum(input_numbers)

  % now work out the mean!
  mean_value = sum_of_input / length_of_input
end
```

The variable `input_numbers` is an input in the **function declaration**

`sum_of_inputs` is created by the function

Whatever is assigned to `mean_value` is produced as an output

---

### The end statement

Indicates the **end** of the function

---

## What's the difference between functions and scripts?

A fundamental difference between **scripts** and **functions** is the
**environment** in which they're executed

**Scripts**

- Scripts have access to **any** variables in the **workspace**

- **All** variables created in a script are created in the **workspace**

**Functions**

- Functions only have access to variables that were passed in as **inputs**

- Only **outputs** from functions will be created in the **workspace**

This is the idea of **variable scope**

---

### Variable scope

**Scripts**

```matlab
% example script saved as example1.m
x = 1;
disp(x); % print out the value of x
```

```matlab
>> x = 99;
>> disp(x); % prints out: 99
x =
    99
    
>> example1; % prints out: 1
     1
     
>> disp(x); % prints out: 1 
x =
     1
```

The value of `x` is set inside the script `example1`, and that change persists
after the script has finished running.

---

### Variable scope

**Functions**
```matlab
% example function saved as example2.m
function example2
  x = 1;
  disp(x); % print out the value of x
end
```

```matlab
>> x = 99;
>> disp(x); % prints out: 99
x =
    99
>> example2; % prints out: 1
    1
>> disp(x); % prints out: 99 
x =
    99
```

The value of `x` is set inside the **function** `example2` and this change
**does not** persist

The `x` **inside** the **function** and the `x` **outside** the **function** are
different `x`s

_Note that this is different to how variable scope works in `R` and `Python`._

---

## Saving functions in files

Functions in `Matlab` should be saved in **separate** files.

- Each function that you write should live **inside its own file**.

- The file should have a **file name** that matches the **function name**

For example, the following function…

```matlab
function [output_vars] = my_fun(input_vars)
  % Do something in here
end

```

…should be saved in a file called `my_fun.m`.


---

### Defining multiple functions per file

Sometimes you might want to define functions that are only used **inside other
functions**

**Example**

- You write a function called `mean_ci` that works out a **mean** and a
  **confidence interval**

- Part of working out a confidence interval involves working out a **standard
  error**.

- So, you write another function called `se` which works out a standard error
  that can be used by your `mean_ci` function.


For a case like this, you have two options:

1. Create two functions in **two** files

2. Create two functions in **one** file

---


### <small>Defining multiple functions per file</small>

Storing the two functions in **one** file:

```matlab
% functions stored in file mean_ci.m
function [mean, ci] = mean_ci(input_numbers)
  mean = sum(input_numbers) / length(input_numbers)
  ci = 1.96 * se(input_numbers)
end

function [sem] = se(input_numbers)
  sd = std(input_numbers)
  sem = sd / sqrt(length(input_numbers))
end
```

- **Advantage**: Fewer files, so easier to manage 

- **Disadvantage**: The function `se` can only be called from inside the
  `mean_ci` function:

```matlab
>> se(input_numbers)
Unrecognized function or variable 'se'.
```
